---
title: "Bayes 2 lecture (EMLAR 2022)"
author: "Shravan Vasishth"
date: "4/21/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(brms)
library(bayesplot)
library(cowplot)
library(tidyverse)
```

# Review of Bayes 1 lecture

The main points that I discussed were as follows.

## Bayes' rule allows us to compute/derive the posterior distribution of a
parameter or parameters of interest:

$$f(\mu | y) \propto f(y|\mu)\times f(\mu)$$

## Example: Complications in an operation

I had discussed the example informally/graphically, but here is a more formal 
presentation:

- We are modeling the number of complications that occur in operations.
- Operationalize the occurrence of a complication with 1, no complication 0.
- This suggests a binomial likelihood; 
  - k is the number of successes
  - n is the total number of trials (roughly: independent data points)
  - $theta$ is the probability parameter (the probability of complications)

\begin{equation}
\mathit{Binomial}(k|n,\theta) = 
\binom{n}{k} \theta^{k} (1-\theta)^{n-k}
\end{equation}

In R, this function is the `dbinom` function; in this function, k is called x, n is called size, and prob refers to theta (because otherwise life would be too easy). Once the data are collected, the data are fixed values (constants). One can then see `dbinom` as a function of $\theta$. That is called the likelihood function.

As an example, if we had k=0 complications (successes here), and n=10, then the likelihood function $f(n,k|\theta)$ is:

```{r }
theta<-seq(0,1,by=0.01)
plot(theta,dbinom(x=0,size=10,prob=theta),type="l")
```

Notice that the maximum likelihood estimate is k/n, here 0/10=0. This is the estimate of $\theta$ that yields the maximum point in the likelihood function above.

**Exercise**: Verify graphically that if k=5, n=10, then the maximum point is at $\theta=0.5$.

### Applying Bayes' rule in our simple example

Likelihood: $\binom{n}{k} \theta^{k} (1-\theta)^{n-k}$

Prior on theta is Beta(3,27). 

\begin{equation}
f(\theta|a,b)=  \frac{1}{B(a=3,b=27)} \theta^{a - 1} (1-\theta)^{b-1}   
(\#eq:beta)
\end{equation}

The posterior (up to proportionality): 


\begin{equation}
p(\theta|n,k)\propto  \theta^{k} (1-\theta)^{n-k} \theta^{a - 1} (1-\theta)^{b-1}
(\#eq:beta)
\end{equation}

This gives us the posterior of $\theta \sim Beta(n+a,n+b-k)$.

### Insight: The posterior distribution is a compromise between the prior and the likelihood

[This is an excerpt from our book.]

Let the data be k=80, n=100. This could be a question-response accuracy for example.

Just for the sake of illustration, let's take four different beta priors, each reflecting increasing certainty. 

  - $\mathit{Beta}(a=2,b=2)$
  - $\mathit{Beta}(a=3,b=3)$
  - $\mathit{Beta}(a=6,b=6)$
  - $\mathit{Beta}(a=21,b=21)$

Each prior reflects a belief that $\theta=0.5$, with varying degrees of (un)certainty. Given the general formula we developed above for the beta-binomial case, we just need to plug in the likelihood and the prior to get the posterior:

\begin{equation}
p(\theta | n,k) \propto p(k |n,\theta) p(\theta)
\end{equation}

The four corresponding posterior distributions would be:

\begin{equation}
p(\theta\mid k,n) \propto [\theta^{80} (1-\theta)^{20}] [\theta^{2-1}(1-\theta)^{2-1}] = \theta^{82-1} (1-\theta)^{22-1}
\end{equation}

\begin{equation}
p(\theta\mid k,n) \propto [\theta^{80} (1-\theta)^{20}] [\theta^{3-1}(1-\theta)^{3-1}] = \theta^{83-1} (1-\theta)^{23-1}
\end{equation}

\begin{equation}
p(\theta\mid k,n) \propto [\theta^{80} (1-\theta)^{20}] [\theta^{6-1}(1-\theta)^{6-1}] = \theta^{86-1} (1-\theta)^{26-1}
\end{equation}

\begin{equation}
p(\theta\mid k,n) \propto [\theta^{80} (1-\theta)^{20}] [\theta^{21-1}(1-\theta)^{21-1}] = \theta^{101-1} (1-\theta)^{41-1}
\end{equation}

We can visualize each of these triplets of priors, likelihoods and posteriors; see Figure \@ref(fig:postbetavizvar). 

```{r postbetavizvar, echo=FALSE,fig.cap = "The (scaled) likelihood, prior, and posterior in the beta-binomial conjugate example, for different uncertainties in the prior. The likelihood is scaled to integrate to 1 to make its comparison easier.  "}
k <- 80
n <- 100
## Prior
a <- 4
b <- 4
binom_lh <- function(theta) {
dbinom(x=k, size =n, prob = theta)
}

K <- integrate(f = binom_lh, lower = 0, upper = 1)$value

binom_scaled_lh <- function(theta) 1/K * binom_lh(theta)
  
p_beta <- ggplot(data = tibble(theta = c(0, 1)), aes(theta)) +
  stat_function(
    fun = dbeta,
    args = list(shape1 = a, shape2 = b),
    aes(linetype = "Prior")
  ) +
  ylab("density") +
  stat_function(
    fun = dbeta,
    args = list(shape1 = k + a, shape2 = n - k + b), aes(linetype = "Posterior")
  ) +
  stat_function(
    fun = binom_scaled_lh,
    aes(linetype = "Scaled likelihood")
  ) +
  theme_bw() +
  theme(legend.title = element_blank())
p_beta
```

If you hold the likelihood function constant (the data are constant at $n=100, k=80$ in the above example), the tighter the prior, the greater the extent to which the posterior orients itself towards the prior.  In general, we can say the following about the likelihood-prior-posterior relationship:

- The posterior distribution is a compromise between the prior and the likelihood.
- For a given set of data, the greater the certainty in the prior, the more heavily the posterior will be influenced by the prior mean.
- Conversely, for a given set of data, the greater the *un*certainty in the prior, the more heavily the posterior will be influenced by the likelihood.

$ Example: Fitting a linear mixed model for a planned experiment

Read in and prepare the two data sets:

```{r}
## load example data-set:
gw<-read.table("data/gibsonwu2012data.txt",
               header=TRUE)
## sum-contrast coding of predictor:
gw$so <- ifelse(
  gw$type%in%c("subj-ext"),-1,1)
## subset critical region
gw1<-subset(gw,region=="headnoun")

## load second data-set:
gw2<-read.table("data/gibsonwu2012datarepeat.txt",
                header=TRUE)
gw2$so <- ifelse(
  gw2$condition%in%c("subj-ext"),-1,1)
```
Frequentist analysis:

```{r}
## frequentist analysis:
library(lme4)
m_lmer<-lmer(rt~so + (1+so|subj)+(1+so|item),gw1)
summary(m_lmer)
```

Always visualize the data first:

```{r}
boxplot(rt~so,gw1)
```

Better:

```{r}
p1CN <- ggplot(gw1, aes(x=type, y=rt)) + 
  geom_boxplot() + ggtitle("Chinese RCs")
p1CN<-p1CN+geom_jitter(shape=16, 
                       position=position_jitter(0.2))+
  theme_bw()+#scale_y_continuous(trans='log2')+
  ylab("Reading times (ms)")+
  coord_cartesian(ylim = c(100,7500)) 
p1CN
```

Log-transformed reading times:

```{r}
m_lmerlog<-lmer(log(rt)~so + (1+so|subj)+(1+so|item),gw1)
summary(m_lmerlog)

boxplot(log(rt)~so,gw1)
```

Bayesian analysis:

```{r cache=TRUE,message=FALSE,warning=FALSE,results='hide'}
priors <- c(set_prior("normal(6, 1.5)", class = "Intercept"),
            set_prior("normal(0, .05)", class = "b", 
                      coef = "so"),
            set_prior("normal(0, 1)", class = "sd"),
            set_prior("normal(0, 1)", class = "sigma"),
            set_prior("lkj(2)", class = "cor"))

m_gw<-brm(rt~so + (1+so|subj) + (1+so|item),gw1,family=lognormal(),
          prior=priors)
summary(m_gw)
```

Posterior predictive check:

```{r}
pp_check(m_gw)
```

Summarize the effect of interest (in ms), and summarize individual-level variation:

```{r}
## graphical visualization:
postgw<-posterior_samples(m_gw)
## extract posteriors:
alpha<-postgw$b_Intercept
beta<-postgw$b_so
cor<-posterior_samples(m_gw,"^cor")
sd<-posterior_samples(m_gw,"^sd")
sigma<-posterior_samples(m_gw,"sigma")

## subject level effects:
subj_re<-posterior_samples(m_gw,"^r_subj")
item_re<-posterior_samples(m_gw,"^r_item")
```

Mean effect in ms:

```{r}
## mean effect on ms scale:
meandiff<- exp(alpha + beta) - exp(alpha - beta)
mean(meandiff)
round(quantile(meandiff,prob=c(0.025,0.975)),0)

## mean effect:
hist(meandiff,freq=FALSE,
     main="Mean OR vs SR processing cost",
     xlab=expression(exp(alpha + beta)- exp(alpha - beta)))
```

Individual differences:

```{r}
## individual level estimates:
nsubj<-37
subjdiff<-matrix(rep(NA,nsubj*4000),nrow=nsubj)
for(i in 1:nsubj){
  subjdiff[i,]<-exp(alpha + subj_re[,i]  + (beta+subj_re[,i+nsubj])) - 
    exp(alpha + subj_re[,i] - 
          (beta+subj_re[,i+nsubj]))
}

subjdiff<-t(subjdiff)

subjdiff<-as.data.frame(subjdiff)
colnames(subjdiff)<-paste("s",c(1:nsubj),sep="")
mns <- colMeans(subjdiff)
subjdiff<-subjdiff[,order(mns)]
mcmc_areas(subjdiff)
```

Shrinkage in action:

```{r}
## lmer
m_lmer<-lmer(log(rt)~so + (1+so|subj),gw1)
summary(m_lmer)


alphalmer<-coef(m_lmer)$subj[,1]
betalmer<-coef(m_lmer)$subj[,2]
lmerest<-exp(alphalmer+betalmer)-exp(alphalmer-betalmer)


## lmList:
m_lmlist<-lmList(log(rt)~so|subj,gw1)
summary(m_lmlist)

alphalmlist<-coef(m_lmlist)[1]
betalmlist<-coef(m_lmlist)[2]
lmlistest<-exp(alphalmlist+betalmlist)-exp(alphalmlist-betalmlist)

comparison<-data.frame(subj=factor(1:37),brm=mns,lmlist=lmlistest,lmer=lmerest)
colnames(comparison)[c(3,4)]<-c("lmlist","lmer")
```

```{r}
plot(lmlist~subj,comparison,ylab="estimate",main="lmlist vs Bayes\n shrinkage in action")
points(brm~subj,pch=3,comparison)
arrows(x0=1:37,y0=comparison$lmlist,x1=1:37,y1=comparison$brm,
       angle=45,code=2,length=0.1,col="red")
```

```{r}
plot(lmlist~subj,comparison,ylab="estimate",main="lmlist vs lmer\n shrinkage in action")
points(lmer~subj,pch=3,comparison)
arrows(x0=1:37,y0=comparison$lmlist,x1=1:37,y1=comparison$lmer,
       angle=45,code=2,length=0.1,col="red")

plot(lmer~subj,comparison,ylab="estimate",main="lmer vs Bayes\n regularization in action")
points(brm~subj,pch=3,comparison)
arrows(x0=1:37,y0=comparison$lmer,x1=1:37,y1=comparison$brm,
       angle=45,code=2,length=0.1,col="red")
```


# Bayes factors analysis

```{r cache=TRUE,message=FALSE,warning=FALSE,results='hide'}
## Bayes factor analysis:
m_gw<-brm(rt~so + (1+so|subj) + (1+so|item),gw1,family=lognormal(),
          prior=priors,warmup=5000,iter=20000,
          save_pars=save_pars(all=TRUE))
summary(m_gw)

priors0 <- c(set_prior("normal(6, 1.5)", class = "Intercept"),
             set_prior("normal(0, 1)", class = "sd"),
             set_prior("normal(0, 1)", class = "sigma"),
             set_prior("lkj(2)", class = "cor"))

m_gw0<-brm(rt~1 + (1+so|subj),gw1,family=lognormal(),
           prior=priors0,,warmup=5000,iter=20000,
           save_pars=save_pars(all=TRUE))

```


Compute the BF several times to check that it is stable:

```{r}
bayes_factor(m_gw,m_gw0)
bayes_factor(m_gw,m_gw0)
```

The Bayes factor is sensitive to the prior, so a sensitivity analysis is a must. Neer report just one Bayes factor with a (vague) prior on the target parameter. See:

Daniel J. Schad, Bruno Nicenboim, Paul-Christian Bürkner, Michael Betancourt, and Shravan Vasishth. Workflow Techniques for the Robust Use of Bayes Factors. Psychological Methods, 2022.
https://osf.io/y354c/


